---
title: "QTL effect simulation in multiple cross population"
author:
  - Vincent Garin^[CIRAD, vincent.garin@cirad.fr]
  - Capucine Mayoud^[Université de Montpellier]
date: "2024-10-02"
output: rmarkdown::html_vignette
bibliography: bibliography.bib
link-citations: yes
vignette: >
  %\VignetteIndexEntry{QTL effect simulation in multiple cross population}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

<style type="text/css">
  body{font-size: 14pt;}
  code.r{font-size: 14pt;}
  pre {font-size: 14pt}
</style>

## Simulation of genotype data

Simulation of genotype data using the 'mpp_sim_cross()' function which is a wrapper function for the 'simcross' package

```{r sim_geno}
library(mppSim)
library(GPfunctional)

data("geno_par")
rownames(geno_par) <- paste0('P', 1:nrow(geno_par))

data("EUNAM_map")
rownames(map) <- map[, 1]

# NAM crossing scheme with 9 parents
cross_scheme <- cross_scheme_NAM(9)

geno <- sim_mpp_cross(geno_par = geno_par, map = map,
                      cross_scheme = cross_scheme,
                      n_ind_cr = rep(100, nrow(cross_scheme)))

```

Extension:

1. simulation of different crossing scheme (NAM, factorial design, diallel)
2. simulation of different type of cross (F2, BCxSy, RIL, DH)

## Simulation of phenotypes

Simulation of phenotype data based on the genotype data.

```{r sim_pheno}
geno_c1 <- geno$geno_num_IBD[1:100, ]
colnames(map)[1] <- "mk.names"

d <- GPF_sim_pheno(X = geno_c1, map = map, n_QTL = 1, prop_f = 1, h2f = 1,
                   corr_f = 1/sqrt(2))

var(d$d_y$y_f)
```

Check: get the last version of the function used by Capucine. Parameters and functionalities: Single cross, multiple cross, ...

## Comparison realized pheno/geno variance with expectation

## Check Linkage disequilibrium pattern

## Compare simulated (genotypic) variance with mixed model estimation

## Simulation and utilisation of data for genomic prediction

## Simulation and utilisation of data for QTL detection 
