---
title: "Theoretical elements of simulation in multiple cross population"
author:
  - Vincent Garin^[CIRAD, vincent.garin@cirad.fr]
  - Capucine Mayoud^[Université de Montpellier]
date: "2024-10-02"
output: rmarkdown::html_vignette
bibliography: bibliography.bib
link-citations: yes
vignette: >
  %\VignetteIndexEntry{Theoretical elements of simulation in multiple cross population}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

<style type="text/css">
  body{font-size: 14pt;}
  code.r{font-size: 14pt;}
  pre {font-size: 14pt}
</style>

## Single cross F2 HWE population

### Marker score variance

Let us assume a bi-allelic SNP marker $X$ where the two alleles frequencies are defined as such

-   $p$ : major ($A$) allele frequency

-   $q$ : minor ($a$) allele frequency

-   $p + q = 1$

Assuming Hardy-Weinberg equilibrium, we have the following genotype frequencies

$$
\begin{align}
X & = \begin{cases}
AA \quad & 0 \quad \text{with a probability} \quad p^2\\
Aa \quad & 1 \quad \text{with a probability} \quad 2pq\\
aa \quad & 2 \quad \text{with a probability} \quad q^2 \\
\end{cases}\\
\end{align}
$$ 

We can calculate the variance of such a random variable using the general following expression of the variance

$$V[X] = E[X^2] - E[X]^2$$
The expectation is expressed as such

$$
\begin{align}
E[X] & = 0 * p^2 + 1 * 2pq + 2 * q^2 \\
     & = 2pq + 2q^2 = 2q(p+q)\\
     & = 2q\\
\end{align}
$$
And the expectation of the squared variable can be developped like that

$$
\begin{align}
E[X^2] & = 0^2 * p^2 + 1^2 * 2pq + 2^2 * q^2 \\
     & = 2pq + 4q^2
\end{align}
$$

By combining those expression we get the expression for the marker score variance

$$
\begin{align}
V[X] & = 2pq + 4q^2 - (2p)^2 \\
     & = 2pq
\end{align}
$$


### Single QTL variance

Let us now derive the genetic variance of the previous marker assuming that this marker is a quantitative trait locus with an additive effect $a$ associated to it. In that case we have the following phenotypes

$$
\begin{align}
G(AA) & = -a \quad \text{with probability} \quad p^2\\
G(AB) & = d  \quad \text{with probability} \quad 2 p q\\
G(BB) & = a  \quad \text{with probability} \quad q^2\\
\end{align}
$$

For simplicity we will assume no dominance effect ($d = 0$).

The expected phenotypic value $E[y]$ of such a single locus model is

$$
\begin{align}
E[y] & = p^2 * a - q^2 * a \\
     & = a*(p^2 - q^2) \\
     & = a*(p+q)(p-q) \\
     & = a*(p-q)
\end{align}
$$
The expected squared phenotypic value $E[y^2]$ is

$$
\begin{align}
E[y^2] & = p^2 * a^2 + q^2 * a^2 \\
       & = a^2*(p^2 + q^2)
\end{align}
$$

Therefore the variance $V[y] = E[y^2] - E[y]^2$

$$
\begin{align}
\sigma_G^2 & =  a^2*(p^2 - q^2) - [a*(p-q)]^2\\
           & =  a^2*(p^2 + q^2) - [a^2*(p^2 - 2pq + q^2)]\\
           & =  2pq a^2\\
\end{align}
$$

### Expression of the additive effect given $\sigma_G^2$

From the last expression we can derive an expression for the additive effect given the additive genetic variance that could be used to simulate effect corresponding to a certain variance (We still assume $d = 0$).

$$
\begin{align}
\sigma_{G}^2 & = 2pq a^2 \\
a^2 & = \frac{\sigma_{G}^2}{2pq}\\
a & = \pm\sqrt{\frac{\sigma_{G}^2}{2pq}}\\
\end{align}
$$

### Simulated additive effect to match expected and realized variance

We must sample QTL effects so that the realized numerical variance match the simulated expected variance.

Let us define $\sigma_P^2$ the realized (numerical) phenotypic variance. This term is defined as the product of the QTL marker score and its aditive effect. We assume a single QTL that account for the whole genetic effect.

$$
\begin{align}
P & = X * a \\
V(P) = \sigma_{P}^2 & = V(X * a)\\
       \sigma_{P}^2 & = V(X) * a^2\\
                    & = \sigma_{X}^2 * \frac{\sigma_{G}^2}{2pq}\\
                    & = \sigma_{X}^2 * \frac{\sigma_{G}^2}{2pq}\\
                    & = \sigma_{X}^2 * \frac{1}{2pq} * \sigma_{G}^2\\
\end{align}
$$

We would like that the phenotypic variance is equivalent to the genotypic variance $\sigma_P^2 = \sigma_G^2$. We can distinghuish two situations. If the marker score $X$ is not standardized ($\sigma_{X}^2 = 2pq$)

$$
\sigma_{P}^2 = 2pq * \frac{1}{2pq} * \sigma_{G}^2 = \sigma_{G}^2 
$$
In that case $\sigma_{P}^2 = \sigma_{G}^2$. No need for correction of the simulated QTL additive effect (to be checked).

If the marker score is standardized ($\sigma_{X}^2 = 1$), then

$$
\begin{align}
\sigma_{P}^2 & = 1 * \frac{1}{2pq} * \sigma_{G}^2 \\
             & = \frac{1}{2pq} \sigma_{G}^2 \\
             & = \frac{1}{2*0.5*0.5} \sigma_{G}^2 \quad \text{assuming} \quad p=q=0.5 \\
             & = 2 \sigma_{G}^2 \\
\end{align}
$$

In that case, we need to add a factor in front of the simulated additive effect

$$
\begin{align}
a & = \pm\sqrt{\frac{1}{2}}\sqrt{\frac{\sigma_{G}^2}{2pq}}\\
  &  = \pm\sqrt{\frac{\sigma_{G}^2}{4pq}}\\
\end{align}
$$

to  get

$$
\begin{align}
\sigma_{P}^2 & = \sigma_{X}^2 * a^2\\
             & = 1 * \frac{1}{2} * \frac{1}{2pq} * \sigma_{G}^2\\
             & = 1 * \frac{1}{2} * 2 * \sigma_{G}^2\\
\sigma_{P}^2 & =  \sigma_{G}^2\\
\end{align}
$$


```{r F2 single QTL variance}
sim_pheno <- c()
```

### Extension of the formula for multiple QTL

If we assume that the QTL only account for a fraction of the total variance $k_i$, we can modify the formula like that to give an expression $\beta_i$ for the $i^{th}$ QTL effect.


$$
\begin{align}
\beta_i &  = \pm\sqrt{\frac{1}{2}}\sqrt{\frac{k_i * \sigma_{G}^2}{2pq}}\\
        &  = \pm\sqrt{\frac{k_i * \sigma_{G}^2}{4pq}}\\
\end{align}
$$

This is similar to the expression proposed by @mollandin_accounting_2022 to simulate QTL effect using Cattle SNP data

$$
\beta_i = \pm(\frac{1}{2})\sqrt{\frac{k_i*\sigma_{G}^2}{2p_i(1-p_i)}}
$$


There correction factor $\frac{1}{2}$ is different from the $\sqrt{\frac{1}{2}}$ use in the $F_2$ HWE setting.


```{r F2 multiple QTL variance}
sim_pheno <- c()
```

## Multiple cross F2 HWE population

We can extend the previous reasoning and derivation to the multiple crosses ($n_c \ge 2$).

### Marker score variance - two crosses ($n_c = 2$)

We start to develop the formula for a situation with two crosses.

Cross 1

$$
\begin{align}
X & = \begin{cases}
0 & \text{avec une probabilité} \quad p_1^2\\
1 & \text{avec une probabilité} \quad 2p_1q_1\\
2 & \text{avec une probabilité} \quad q_1^2 \\
\end{cases}\\
\end{align}
$$ 

Cross 2

$$
\begin{align}
X & = \begin{cases}
0 & \text{avec une probabilité} \quad p_2^2\\
1 & \text{avec une probabilité} \quad 2p_2q_2\\
2 & \text{avec une probabilité} \quad q_2^2 \\
\end{cases}\\
\end{align}
$$ 


To derive the variance of a random variable distributed over multiple-subspace (crosses), we can use the conditional variance formula for a disjoint partition of the probabilistic space (cross) $C_1, C_2, ..., C_n$

$$
\begin{align}
V[X] & = \sum_{i=1}^n V[X|C_i]*P(C_i) \\
     & + \sum_{i=1}^n E[X|C_i]^2 * (1-P(C_i)P(C_i)) \quad \\
     & -2\sum_{i=2}^n \sum_{j=1}^{i-1}  [E[X|C_i]*P(C_i)] * [E[X|C_j]*P(C_j)] \\
\end{align}
$$

In the case of a two crosses we have

$$
\begin{align}
V[X] & = V[X|C_1]*\phi_1 + V[X|C_2]*\phi_2 \\
     & + E[X|C_1]^2 * \delta_1 + E[X|C_2]^2 * \delta_2 \quad\\
     & - 2*[E[X|C_1]*\phi_1] * [E[X|C_2]*\phi_2] \\
\end{align}
$$


Where $\phi_1$ and $\phi_2$ represent the probability to be in cross 1 and 2,  respectively. $\delta_i = (\phi_i - 1)\phi_i$.

So we need to estimate the following components

$E[X|C_1]$

$E[X|C_2]$

$E[X^2|C_1]$

$E[X^2|C_2]$

$V[X|C_1]$

$V[X|C_2]$

Development of the different terms of the expressions

$$
\begin{align}
E[X|C_1] & = 0 * p_1^2 + 1 * 2p_1q_1 + 2 * q_1^2\\
         & = 2p_1q_1 + 2q_1^2\\
         & = 2q_1(p_1 + q_1)\\
         & = 2q_1 * 1 \\
         & = 2q_1
\end{align}
$$

And,

$$
\begin{align}
(E[X|C_1])^2 & = (2q_1)^2\\
             & = 4q_1^2
\end{align}
$$

And,

$$
\begin{align}
E[X^2|C_1] & = 0^2 * p_1^2 + 1^2 * 2p_1q_1 + 2^2 * q_1^2\\
           & = 2p_1q_1 + 4q_1^2\\
\end{align}
$$

Therefore, the within cross conditional variance expression

$$
\begin{align}
V[X|C_1] & = E[X^2|C_1] - (E[X|C_1])^2 \\
         & = (2p_1q_1 + 4q_1^2) - 4q_1^2\\
         & = 2p_1q_1
\end{align}
$$


Similarly for the 2nd cross:

$$
\begin{align}
E[X|C_2] & = 2q_2 \\
\end{align}
$$

And,

$$
\begin{align}
(E[X|C_2])^2 & = 4q_2^2 \\
\end{align}
$$

And,

$$
\begin{align}
E[X^2|C_2] & = 2p_2q_2 + 4q_2^2\\
\end{align}
$$

And the variance,

$$
\begin{align}
V[X|C_2] & = 2p_2q_2
\end{align}
$$


Finaly, we can susstitute all those terms in the general variance formula :

$$
\begin{align}
V[X] & = V[X|C_1]*\phi_1 + V[X|C_2]*\phi_2 \\
     & + E[X|C_1]^2 * \delta_1 + E[X|C_2]^2 * \delta_2 \quad\\
     & - 2*[E[X|C_1]*\phi_1] * [E[X|C_2]*\phi_2] \\
\end{align}
$$

We get

$$
\begin{align}
V[X] & = 2p_1q_1*\phi_1 + 2p_2q_2 * \phi_2 \\
     & + 4q_1^2 * \delta_1 + 4q_2^2 * \delta_2\\
     & - 2 * (2q_1) * \phi_1 * (2q_2) * \phi_2 \\
\end{align}
$$

After simplification

$$
\begin{align}
V[X] & = 2p_1q_1*\phi_1 + 4q_1^2 * \delta_1\\
     & + 2p_2q_2 * \phi_2 + 4q_2^2 * \delta_2\\
     & - 8 * q_1 * q_2 * \phi_1 * \phi_2 \\
\end{align}
$$

Assuming that $p = q = 0.5$, $\phi = 0.5$, we have $\delta = 0.25$, and :

$$
\begin {align}
V[X] & = 2 * 0.5^3 + 4 * 0.5^2 * 0.25 \\
     & + 2 * 0.5^3 + 4 * 0.5^2 * 0.25 \\
     & - 8 * 0.5^4 \\
V[X] & = 0.25 + 0.25 + 0.25 + 0.25 - 0.5 \\
     & = 0.5 \\
\end {align}
$$

### Marker score variance - generalisation for $n_c > 2$

Pour $i \neq j$ 
$$
V(X) = \sum_{i = 1}^{nc} X_i - (\sum_{i = 1}^{nc}\sum_{j = i + 1}^{nc} X_{ij} ) ]
$$

$X_i = 2p_iq_i\phi_i + 4q_i^2\delta_i$ 

$X_{ij} = 8q_iq_j\phi_i\phi_j$

On veut démontrer que la variance du score du marqueur est bien toujours
égale à $\frac{1}{2}$, quelque soit le nombre de croisement (avec
$p = q = 0.5$, et $\phi = 1/nc$, d'où $\delta = \phi(1-\phi)$).

Tout d'abord, on substitue les valeurs dans nos constantes: 
$$
\begin {align}
X_i & = 2 * \frac{1}{2} * \frac{1}{2} * \frac{1}{nc} + 4 * (\frac{1}{2})^2 * \frac{1}{nc} (1 - \frac{1}{nc}) \\
    & = \frac{1}{2nc} + \frac{1}{nc} - \frac{1}{nc^2}\\
    & = \frac{3}{2nc} - \frac{1}{nc^2}
\end {align} 
$$
Ensuite, 
$$
\begin {align}
X_{ij} & = 8 * \frac{1}{2} * \frac{1}{2} * \frac{1}{nc} * \frac{1}{nc}\\
       & = 8 * \frac{1}{4} * \frac{1}{nc^2} \\
       & = \frac{2}{nc^2}
\end {align} 
$$

On réinjecte dans l'expression initale :

Première partie de la somme : 
$$ 
\begin {align}
\sum_{i = 1}^{nc} X_i &  = \sum_{i = 1}^{nc} (\frac{3}{2nc} - \frac{1}{nc^2})\\
                      & = nc *(\frac{3}{2nc} - \frac{1}{nc^2}) \\
                      & = \frac{3}{2} - \frac{1}{nc} \\
\end {align} 
$$

Deuxième partie de la somme : 

$$ 
\begin {align}
\sum_{i = 1}^{nc}\sum_{j = i + 1}^{nc} X_{ij} & = \sum_{i = 1}^{nc}\sum_{j = i + 1}^{nc} (\frac{2}{nc^2}) \\
                                              & = \binom{nc}{2} * (\frac{2}{nc^2})\\
                                              & = \frac{nc(nc-1)}{nc} * \frac{2}{nc^2} \\
                                              & = \frac{nc-1}{nc}\\
\end {align} 
$$

On remplace dans l'expression initiale :

$$
\begin{align}
V(X) & = \sum_{i = 1}^{nc} X_i - (\sum_{i = 1}^{nc}\sum_{j = i + 1}^{nc} X_{ij} ) ] \\
     & = \frac{3}{2} - \frac{1}{nc}  - \frac{nc-1}{nc} \\
     & = \frac{3nc - 2 - 2(nc-1)}{2nc}\\
     & = \frac{3nc - 2 - 2nc + 2}{2nc}\\
     & = \frac{nc}{2nc}\\
     & = \frac{1}{2}
\end {align} 
$$

On a donc bien ce qu'on voulait démontrer, la variance du score des
marqueurs est bien égale à $\frac{1}{2}$ quelque soit le nombre de
croisement.

Therefore, we notice that in the HWE situation the marker score variance is consistent over crosses and it is always equal to $\frac{1}{2}$ (single crosses: $V[X] = 2pq = \frac{1}{2}$)

### Single QTL variance in multiple crosses - two crosses ($n_c = 2$)

Il est possible d'étendre la formule à une situation avec plusieurs
croisements. Pour cela, commençons par une situation avec deux
croisements par souci de simplicité.

On suppose que le marqueur suivant dépend uniquement d'effets additifs
(donc l'hétérozygote correspond à la moitié de l'homozygote). On
considère également que l'on se place dans une population NAM où le
parent central (référence) a un allèle valant 0.

Cross 1

$$
\begin{align}
G(AA) & = \beta_{A} = 0 \quad \text{avec une probabilité} \quad p_1^2\\
G(AB) & = \beta_{AB} = \frac{1}{2}b \quad \text{avec une probabilité} \quad 2 p_1 q_1\\
G(BB) & = \beta_{B} = b \quad \text{avec une probabilité} \quad q_1^2\\
\end{align}
$$ 

Cross 2

$$
\begin{align}
G(AA) & = \beta_{A} = 0 \quad \text{avec une probabilité} \quad p_2^2\\
G(AC) & = \beta_{AC} = \frac{1}{2}c \quad \text{avec une probabilité} \quad 2 p_2 q_2\\
G(CC) & = \beta_{C} = c \quad \text{avec une probabilité} \quad q_2^2\\
\end{align}
$$

Dans ce cas, la formule de la variance conditionnelle pour une partition
disjointe dans un espace probabiliste (croisement) $C_1, C_2, ..., C_n$
est utilisée.

$$
\begin{align}
V[X] & = \sum_{i=1}^n V[X|C_i]*P(C_i) \\
     & + \sum_{i=1}^n E[X|C_i]^2 * (1-P(C_i)P(C_i)) \quad \\
     & -2\sum_{i=2}^n \sum_{j=1}^{i-1}  [E[X|C_i]*P(C_i)] * [E[X|C_j]*P(C_j)] \\
\end{align}
$$

Dans un cas avec deux croisements, on a :

$$
\begin{align}
V[X] & = V[X|C_1]*\phi_1 + V[X|C_2]*\phi_2 \\
     & + E[X|C_1]^2 * \delta_1 + E[X|C_2]^2 * \delta_2 \quad\\
     & - 2*[E[X|C_1]*\phi_1] * [E[X|C_2]*\phi_2] \\
\end{align}
$$

Où $\phi_1$ et $\phi_2$ sont les probabilités d'être dans le croisement
1 ou 2 respectivement, et $\delta_i = (\phi_i - 1)\phi_i$.

Il faut estimer les composantes suivantes :

$E[X|C_1]$

$E[X|C_2]$

$E[X^2|C_1]$

$E[X^2|C_2]$

$V[X|C_1]$

$V[X|C_2]$

Development of the different expressions

$$
\begin{align}
E[X|C_1] & = (p_1^2 * \beta_{A}) + \beta_{AB} * 2 p_1 q_1  + (q_1^2 * \beta_{B}) \\
         & = (p_1^2 * 0) + \frac{1}{2}b * 2 p_1 q_1  + (q_1^2 * b) \\
         & = p_1 q_1 b  + q_1^2 * b \\
         & = b*(p_1 q_1 + q_1^2)\\
         & = b*(q_1 * (p_1 + q_1))\\
         & = b*q_1\\
\end{align}
$$

$$
\begin{align}
E[X|C_2] & = (p_2^2 * \beta_{A}) +  2 p_2 q_2 * \beta_{AC}  + (q_2^2 * \beta_{C}) \\
         & = (p_2^2 * 0) + \frac{1}{2}c * 2 p_2 q_2  + (q_2^2 * c) \\
         & = p_2 q_2 c  + q_2^2 * b \\
         & = c*(p_2 q_2 + q_2^2)\\
         & = c*(q_2 * (p_2 + q_2))\\
         & = c*q_2\\
\end{align}
$$

Calculs intermédiaires :

$$
\begin{align}
[E[X|C_1]]^2 & = (b*q_1)^2\\
           & = b^2 * q_1^2\\
\end{align}
$$

$$
\begin{align}
[E[X|C_2]]^2 & = (c*q_2)^2 \\
             & = c^2 * q_2^2 \\
\end{align}
$$

Suite développement expressions :

$$
\begin{align}
E[X^2|C_1] & = (p_1^2 * \beta_{A}^2) + \beta_{AB}^2 * 2 p_1 q_1  + (q_1^2 * \beta_{B}^2) \\
         & = (p_1^2 * 0) + \frac{1}{4}b^2 * 2 p_1 q_1  + (q_1^2 * b^2) \\
         & = \frac{1}{2}p_1 q_1 * b^2  + q_1^2 * b^2 \\
         & = b^2 * ( \frac{1}{2} p_1 q_1 + q_1^2)\\
         & = b^2 * ( \frac{1}{2} (1-q_1) q_1 + q_1^2)\\
         & = b^2 * ( \frac{1}{2} (q_1-q_1^2) + q_1^2)\\
         & = b^2 * [ \frac{1}{2} q_1 - \frac{1}{2}q_1^2 + q_1^2]\\
         & = b^2 * [ \frac{1}{2} q_1 + \frac{1}{2}q_1^2]\\
         & =  \frac{1}{2} b^2  (q_1 + q_1^2)
\end{align}
$$

$$
\begin{align}
E[X^2|C_2] & = (p_2^2 * \beta_{A}^2) + \beta_{AC}^2 * 2 p_2 q_2  + (q_2^2 * \beta_{C}^2) \\
         & = (p_2^2 * 0) + \frac{1}{4}c^2 * 2 p_2 q_2  + (q_2^2 * c^2) \\
         & = \frac{1}{2}p_2 q_2 * c^2  + q_2^2 * c^2 \\
         & = ...\\
         & =  \frac{1}{2} c^2  (q_2 + q_2^2)
\end{align}
$$

Maintenant, on calcule les variances. Pour cela, on utilise :

$$
V[X|C_i] = E[X^2|C_i] - E[X|C_i]^2
$$

D'où :


$$
\begin{align}
V[X|C_1] & = E[X^2|C_1] - E[X|C_1]^2 \\
         & = \frac{1}{2} b^2  (q_1 + q_1^2)  - b^2q_1^2\\
         & = \frac{1}{2} b^2  q_1 + \frac{1}{2} b^2 q_1^2  - b^2q_1^2\\
         & = \frac{1}{2} b^2  q_1 - \frac{1}{2} b^2 q_1^2 \\
         & = \frac{1}{2} b^2  (q_1 - q_1^2) \\
         & = \frac{1}{2} b^2  (q_1 (1 - q_1)) \\
         & = \frac{1}{2} b^2  q_1p_1 \\
\end{align}
$$ 

Et,

$$
V[X|C_2] = \frac{1}{2} c^2  q_2p_2
$$ 
On remplace dans la formule de la variance :

D'où, 

$$
\begin{align}
V[X] & = \frac{1}{2} b^2  q_1p_1 * \phi_1 + \frac{1}{2} c^2  q_2p_2 * \phi_2 \\
     & + b^2 * q_1^2 * \delta_1 + c^2 * q_2^2 * \delta_2 \\
     & - 2*[(bq_1)*\phi_1] * [(cq_2)*\phi_2] \\
\end{align}
$$

On simplifie en regroupant les termes par croisements et par effets
$b^2q_1$ ($c^2q_2$) :

$$
\begin{align}
V[X] & = \frac{1}{2} b^2  q_1p_1 * \phi_1 + b^2 * q_1^2 * \delta_1 \\
     & + \frac{1}{2} c^2  q_2p_2 * \phi_2 + c^2 * q_2^2 * \delta_2 \\
     & - 2*[(bc)*(q_1q_2)*(\phi_1 \phi_2)]  \\
     & \\
\end {align}
$$

D'où : 

$$
\begin {align}
V[X] & = b^2 * (\frac{1}{2} p_1 q_1 \phi_1 + q_1^2 \delta_1) \\
     & + c^2 * (\frac{1}{2} p_2 q_2 \phi_2 + q_2^2 \delta_2) \\
     & - 2bc*[(q_1q_2)*(\phi_1 \phi_2)]  \\
\end{align} 
$$

On pose :

$$C_1 = (\frac{1}{2} q_1p_1 \phi_1) + q_1^2 \delta_1$$

$$C_2 = (\frac{1}{2} q_2p_2 \phi_2) + q_2^2 \delta_2$$

$$C_{12} = 2*(q_1q_2)*(\phi_1 \phi_2)$$

Ainsi,

$$
V[X] = b^2 * C_1 + c^2 * C_2 - bc*C_{12} 
$$

### Expression of the additive effect given $\sigma_G^2$


Ensuite, on cherche à exprimer l'effet d'un QTL à partir de l'expression de sa variance (puisque qu'il n'y a pas de dominance
$V[X] = \sigma^2_G = \sigma^2_A$:

#### Effet constant : b = c

$$
V[X] = b^2 * [C_1 + C_2 - C_{12}] 
$$

D'où, 

$$
\begin{align}
\sigma_G^2 & = b^2 * [C_1 + C_2 - C_{12}] \\
b^2  & = \frac{\sigma_G^2}{[C_1 + C_2 - C_{12}]} \\
b  & = \pm\sqrt{ \frac{\sigma_G^2}{[C_1 + C_2 - C_{12}]}} \\
\end{align}
$$

#### Effet en série allélique : c = sb

$$
\begin{align}
\sigma_{G}^2 & = b^2 * [C_1 + s^2 C_2 - sC_{12}] \\
\end{align}
$$ 

$$
\begin{align}
b^2  & = \frac{1}{4} \frac{\sigma_G^2}{[C_1 + s^2 C_2 - sC_{12}]} \\
b  & = \pm \frac{1}{2} \sqrt{ \frac{(k)*\sigma_G^2}{[C_1 + s^2 C_2 - sC_{12}]}} \\
\end{align}
$$ 

$c = s^2b$

#### Pour 3 croisements :

On note $d$ l'effet du 3e croisement :

Croisement 1 :

$E(X|C_1) = bq_1$
$E(X^2|C_1) = \frac{1}{2}b^2(q_1+q_1^2)$
$V(X|C_1) = \frac{1}{2} b^2q_1p_1$

Croisement 2 :

$E(X|C_2) = cq_2$
$E(X^2|C_2) = \frac{1}{2}c^2(q_2+q_2^2)$
$V(X|C_2) = \frac{1}{2} c^2q_2p_2$


Croisement 3 :

$E(X|C_3) = dq_3$
$E(X^2|C_3) = \frac{1}{2}d^3(q_3+q_3^2)$
$V(X|C_3) = \frac{1}{2} d^2q_3p_3$

Donc on peut calculer la variance avec ces éléments, puis simplifier :


$V(X) = b^2C_1 + c^2C_2 + d^2C_3 - bc C_{12} + bd C_{13} + cd C_{23})$

Avec

$$C_1 = (\frac{1}{2} q_1p_1 \phi_1) + q_1^2 \delta_1$$

$$C_2 = (\frac{1}{2} q_2p_2 \phi_2) + q_2^2 \delta_2$$

$$C_3 = (\frac{1}{2} q_3p_3 \phi_3) + q_3^2 \delta_3$$

$$C_{12} = 2*(q_1q_2)*(\phi_1 \phi_2)$$
$$C_{13} = 2*(q_1q_3)*(\phi_1 \phi_3)$$
$$C_{23} = 2*(q_2q_3)*(\phi_2 \phi_3)$$

#### Effet constant : b = c = d

$$
V(X) = b^2[ (C_1 + C_2 + C_3) - ( C_{12} + C_{13} + C_{23})]
$$

On peut déduire :

$$
\begin{align}
\sigma_G^2 & = b^2[ (C_1 + C_2 + C_3) - ( C_{12} + C_{13} + C_{23})] \\
b^2  & = \frac{\sigma_G^2}{[ (C_1 + C_2 + C_3) - ( C_{12} + C_{13} + C_{23})]} \\
b  & = \pm\sqrt{ \frac{\sigma_G^2}{[ (C_1 + C_2 + C_3) - ( C_{12} + C_{13} + C_{23})]}} \\
\end{align}
$$

#### Effet en série allélique : c = sb, d = s\^2b

$$
\begin{align}
\sigma_{G}^2 & = b^2 * [ (C_1 + s^2 C_2 + s^4C_{3}) - (s C_{12} + s^2 C_{13} + s^3 C_{23})] \\
\end{align}
$$ 

$$
\begin{align}
b^2  & = \frac{\sigma_G^2}{[(C_1 + s^2 C_2 + s^4C_{3}) - (s C_{12} + s^2 C_{13} + s^3 C_{23})]} \\
b  & = \pm \sqrt{ \frac{(k)*\sigma_G^2}{[ (C_1 + s^2 C_2 + s^4C_{3}) - (s C_{12} + s^2 C_{13} + s^3 C_{23})]}} \\
\end{align}
$$

#### Généralisation pour nc croisements :

##### Effet constant

Pour $i \neq j$ 

$$
V(X) = b^2 [\sum_{i = 1}^{nc} C_i - (\sum_{i = 1}^{nc}\sum_{j = i + 1}^{nc} C_{ij} ) ]
$$ 

Avec,

$C_i = (\frac{1}{2} p_i q_i \phi_i) + (q_i^2 \delta_i)$

$C_{ij} = 2 * (q_iq_j)(\phi_i \phi_j)$

D'où :

$$
b = \pm \frac{1}{2}  \sqrt\frac{\sigma_G^2}{[\sum_{i = 1}^{nc} C_i - (\sum_{i = 1}^{nc}\sum_{j = i + 1}^{nc} C_{ij} ) ]}
$$ 

Le facteur 1/2, voir démonstration plus loin.

##### Effets en série allélique

$c = sb$ , $d = s^2b$ etc...

Pour $i \neq j$ 

$$
V(X) = b^2 [\sum_{i = 1}^{nc} s^{2(i-1)}. C_i - (\sum_{i = 1}^{nc} \sum_{j = i + 1}^{nc} s^{i+j-2}.C_{ij} )]
$$ 

$C_i = (\frac{1}{2} p_i q_i \phi_i) + (q_i^2 \delta_i)$

$C_{ij} = 2 * (q_iq_j)(\phi_i \phi_j)$

D'où :

$$
b = \pm (\frac{1}{2}) \sqrt\frac{\sigma_G^2}{[\sum_{i = 1}^{nc} s^{2(i-1)}. C_i - (\sum_{i = 1}^{nc} \sum_{j = i + 1}^{nc} s^{i+j-2}.C_{ij} )]} 
$$


### Simulated additive effect to match expected and realized variance


As in the single cross situation, we need to add a factor to make sure that the simulated numerical (phenotypic) variance match the expected (genetic) variance


As before, we define $\sigma_P^2$ the realized (numerical) phenotypic variance. This term is defined as the product of the QTL marker score and its aditive effect. We assume a single QTL that account for the whole genetic effect.

$$
\begin{align}
P & = X * b \\
V(P) = \sigma_{P}^2 & = V(X * b)\\
       \sigma_{P}^2 & = V(X) * b^2\\
                    & = \sigma_{X}^2 * f^2 * \frac{\sigma_G^2}{[\sum_{i = 1}^{nc} s^{2(i-1)}. C_i - (\sum_{i = 1}^{nc} \sum_{j = i + 1}^{nc} s^{i+j-2}.C_{ij} )]}\\
\end{align}
$$
Considering the case with non-standardized marker scores. We have shown before that in the $n_c$ cross situation $\sigma_{X}^2 = \frac{1}{2}$. Similarly we can show that the denominator expression $[\sum_{i = 1}^{nc} s^{2(i-1)}. C_i - (\sum_{i = 1}^{nc} \sum_{j = i + 1}^{nc} s^{i+j-2}.C_{ij} )]$ is equal to $\frac{1}{8}$. Therefore, we have


$$
\begin{align}
\sigma_{P}^2 & = \frac{1}{2} * f^2 * \frac{\sigma_G^2}{\frac{1}{8}}\\
\sigma_{P}^2 & = \frac{8}{2} * f^2 * \sigma_G^2\\
\sigma_{P}^2 & = 4 * f^2 * \sigma_G^2\\
\end{align}
$$
So $f^2 = \frac{1}{4}$, and $f = \frac{1}{2}$. The formula is then updated like that

$$
b = \pm \frac{1}{2} \sqrt\frac{\sigma_G^2}{[\sum_{i = 1}^{nc} s^{2(i-1)}. C_i - (\sum_{i = 1}^{nc} \sum_{j = i + 1}^{nc} s^{i+j-2}.C_{ij} )]} 
$$

#### Proof about the denominator value

Comme démontré plus haut, la variance théorique du score du marqueur
$V[X]$ est égale à $\frac{1}{2}$.

En prenant les mêmes considérations : $p=q = 0.5$, et $\phi = 0.5$, d'où
$\delta=0.25$ : La valeur du dénominateur $[C_1 + C_2 -2C_{12}]$ est
égale à :

$$
\begin {align}
C_1 & = 0.5 * q_1 * p_1 * \phi_1 + q_1^2 * \delta_1 \\
    & = 0.5^4 + 0.5^2 * 0.25\\
    & = 0.125\\
\end {align}
$$    
$$
\begin {align}
C_2 & = 0.125\\
\end {align}
$$
$$
\begin {align}
C_{12} & = 2q_1q_2\phi_1\phi_2\\
       & = 2 * 0.5^4 \\
       & = 0.125\\
[C_1 + C_2 -2C_{12}] & = 0.125 + 0.125 - 0.125 \\ 
                     & = 0.125 \\
\end {align}
$$

Donc $[C_1 + C_2 -2C_{12}] = \frac{1}{8}$


##### Généralisation

On suppose que $p = q = 0.5$ et que $\phi = \frac{1}{nc}$, d'où
$\delta = \frac{1}{nc} * (1 - \frac{1}{nc})$

On veut démontrer que quelque soit le nombre de croisement, le
dénominateur
$[\sum_{i = 1}^{nc} C_i - (\sum_{i = 1}^{nc}\sum_{j = i + 1}^{nc} C_{ij} ) ]$
est bien égal à $\frac{1}{8}$.

Tout d'abord, on substitue les valeurs dans nos constantes:

$C_i = (\frac{1}{2} p_i q_i \phi_i) + (q_i^2 \delta_i)$

$$
\begin {align}
C_i & = (\frac{1}{2} p_i q_i \phi_i) + (q_i^2 \delta_i) \\
    & = \frac{1}{2} * \frac{1}{2} *\frac{1}{2} * \frac{1}{nc} + (\frac{1}{2})^2 + \frac{1}{nc} * (1 - \frac{1}{nc}) \\
    & = \frac{1}{8nc} + \frac{1}{4nc} * (1 - \frac{1}{nc})\\
    & = \frac{1}{8nc} + \frac{1}{4nc} * \frac{nc-1}{nc}\\
    & = \frac{1}{8nc} + \frac{nc-1}{4nc^2}\\
\end {align} 
$$

$C_{ij} = 2 * (q_iq_j)(\phi_i \phi_j)$

$$
\begin {align}
C_{ij} & = 2 * (q_iq_j)(\phi_i \phi_j) \\
       & = 2 * \frac{1}{2} * \frac{1}{2} * \frac{1}{nc} *\frac{1}{nc} \\
       & = \frac{1}{2nc^2}\\
\end {align} 
$$

On réinjecte dans l'expression initale :
$[\sum_{i = 1}^{nc} C_i - (\sum_{i = 1}^{nc}\sum_{j = i + 1}^{nc} C_{ij} ) ]$

Première partie de la somme: 
$$
\begin {align}
\sum_{i = 1}^{nc} C_i & = \sum_{i = 1}^{nc} (\frac{1}{8nc} + \frac{nc-1}{4nc^2})\\
                      & = nc.(\frac{1}{8nc} + \frac{nc-1}{4nc^2})\\
                      & = \frac{nc}{8nc} + \frac{nc.(nc-1)}{4nc^2}\\
                      & = \frac{1}{8} + \frac{nc-1}{4nc}\\
\end {align} 
$$

Deuxième partie :

$$
\begin {align}
\sum_{i = 1}^{nc}\sum_{j = i + 1}^{nc} C_{ij}  & = \sum_{i = 1}^{nc}\sum_{j = i + 1}^{nc} (\frac{1}{2nc^2}) \\
                                               & = \frac{1}{2nc^2} * \binom{nc}{2} \\
                                               & = \frac{1}{2nc^2} * \frac{nc(nc-1)}{2}\\
                                               & = \frac{nc - 1}{4nc}
\end {align}
$$

En remplacant dans la somme : 
$$
\begin {align}
[\sum_{i = 1}^{nc} C_i - (\sum_{i = 1}^{nc}\sum_{j = i + 1}^{nc} C_{ij} ) ] & = \frac{1}{8} + \frac{nc-1}{4nc} - \frac{nc - 1}{4nc} \\
                                                                    & = \frac{1}{8}\\
\end {align} 
$$

On retombe donc bien sur ce qu'on voulait démontrer :
$\frac{1}{8}$


```{r MPP F2 single QTL variance}
sim_pheno <- c()
```

### Extension of the formula for multiple QTL

$$
b = \pm \frac{1}{2} \sqrt\frac{k*\sigma_G^2}{[\sum_{i = 1}^{nc} s^{2(i-1)}. C_i - (\sum_{i = 1}^{nc} \sum_{j = i + 1}^{nc} s^{i+j-2}.C_{ij} )]} 
$$

## Multiple cross BCxSy HWE population

## Extension to other population type (e.g. BCxSy)

### Genotype freqencies

We can extend the previous derivation in the case of a different segregation pattern. For example $BC_1F_3$ instead of $F_2$.

We still assume a bi-allelic marker system with $f(A) = p$ and $f(a) = 1 - p = q$

At the BC1 generation we have

$AA \quad \times \quad Aa$

|  | A | a 
|--|----|----| 
|A | AA | Aa
|A | AA | Aa  

|  | A | a 
|--|----|----| 
|A | \(p^2\) | \(pq\)
|A | \(p^2\) | \(pq\)

So for the selfing generation there is one part of the population that is fixed $AA$. This part of the population get a probability $2*p^2$.

The other part $Aa$ with a probability $2pq$ still vary. For example generation BC1F2

|  | A | a 
|--|----|----| 
|A | AA | Aa
|a | Aa | aa  

|  | A | a 
|--|----|----| 
|A | \(p^2\) | \(pq\)
|A | \(pq\) | \(q^2\)

So at that generation the $AA$ part with a $2pq * p^2$ part is fixed. Similarly the $aa$ part with $2pq * q^2$ is also fixed.

At the BC1F3 generation the only part tha vary is the $Aa$ with probability $2pq * 2pq$

|  | A | a 
|--|----|----| 
|A | \(p^2\) | \(pq\)
|A | \(pq\) | \(q^2\)

Therefore, when we sum the different probabilities

$$
\begin{align}
P(AA) & = 2p^2 + p^2 * 2pq + p^2 * (2pq)^2 = 2p^2 + p^2 * (\sum_{i=1}^{ns -1} (2pq)^{ns})\\
P(Aa) & = (2pq) * (2pq) * (2pq) = (2pq)^{ns}\\
P(aa) & = (2pq * (q^2)) +  ((2pq) * (2pq) *(q^2)) = q^2 * (\sum_{i=1}^{ns -1} (2pq)^{ns})\\
\end{align}
$$

### Marker score variance

### Single QTL variance

### Simulated additive effect to match expected and realized variance

```{r BCxSy single QTL variance}
sim_pheno <- c()
```

### Extension of the formula for multiple QTL



## References
